---
title: 'GG606: Assignment 1'
author: "Julia K"
date: "2024-02-05"
output: 
  html_document:
    toc: true
    toc_depth: #########################
    toc_float: 
      collapsed: false
      smooth_scroll: false 
    highlight: kate
    number_sections: false
    keep_md: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##**Assignment 1 - Importing, parsing, and querying data in the wild**


### 1. Loading Things: \

##### 1.1: Loading Packages \
Load in any and all packages required to manipulate, transform, and illustrate 
data given. 
```{r}
#Functional:
library(tidyverse)     #Package to process our data, stylized formation
library(dplyr)         #Sub tidyverse package for data manipulation 
library(magrittr)      #Package to help coding sequencing 
library(janitor)       #Package for 'clean_names()' function
library(readr)         #Package to help parse rectangular data
library(stringr)       #Package to deal w/ NAs and manipulate cols/data
library(lubridate)     #Package for date processing and plotting
library(broom.mixed)   #Package to clean diverse model outputs

#Aesthethics:
library(ggplot2)       #Package to generate plotted data
library(patchwork)     #Package for extensive plotted data configuration 
library(ggtheme)       #Package for extra themes, scales, and geoms for plotted data
library(RColorBrewer)  #Package to colour plots
library(viridisLite)   #Package to colour plots

library(here)          #Package to set working directory via `.Rproj`
```



##### 1.2: Loading Data \
The `.csv` data file given has been loaded onto a Google platform and published as
a URL in order to (hopefully) grant universal user access. However, this becomes defeatist when loading the following code chunk, which requires a redetermined
plot theme function. To ignore the errow message, remove code chunk two. 

```{r Loading Data, inlcude=TRUE}
https://docs.google.com/spreadsheets/d/e/2PACX-1vREAF1nhXX9kzvysMZQqlRIM6XPnlrjDtlmhXb8OsRcHuu2S3SIpI-xQnieAmIlQWfAeMP-IklnCZhl/pubhtml?gid=1389465964&single=true
```


##### 1.3: Loading Select Functions \

We will be applying a pre-determined function that has previously been generated. Here, the function is titled theme_juicy.R, and will be used for any and all figures generated.
```{r Loading Functions, include=TRUE}
source(here("functions", "theme_juicy.R")) #load from 'functions' folder
theme_set(theme_juicy())                   #auto-apply as theme for plots
```

