---
title: 'GG606: Assignment 1'
author: "Julia K"
date: "2024-02-05"
output: 
  html_document:
    toc: true
    toc_depth: #########################
    toc_float: 
      collapsed: false
      smooth_scroll: false 
    highlight: kate
    number_sections: false
    keep_md: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#**Assignment 1 - Importing, parsing, and querying data in the wild**


## 1. Loading Things: \

#### 1.1: Loading Packages \
Load in any and all packages required to manipulate, transform, and illustrate 
data given. 
```{r Loading Packages, include=TRUE}
#Functional:
library(googlesheets4)
library(tidyverse)     #Package to process our data, stylized formation
library(dplyr)         #Sub tidyverse package for data manipulation 
library(vroom)
library(magrittr)      #Package to help coding sequencing 
library(janitor)       #Package for 'clean_names()' function
library(readr)         #Package to help parse rectangular data
library(stringr)       #Package to deal w/ NAs and manipulate cols/data
library(lubridate)     #Package for date processing and plotting
library(broom.mixed)   #Package to clean diverse model outputs
library(fedmatch)
#Aesthetics:
library(ggplot2)       #Package to generate plotted data
library(patchwork)     #Package for extensive plotted data configuration 
library(ggthemes)      #Package for extra themes, scales, and geoms for plotted data
library(RColorBrewer)  #Package to colour plots
library(viridisLite)   #Package to colour plots

library(here)          #Package to set working directory via `.Rproj`
#getwd()               #Function to affirm working directory 
```


________________________________________________________________________________

### Question 1: \
*Read the data in and clean it for analysis, using the readr package functions for reading and parsing data. Provide a few comments on the types of earthquakes and the sources of information of the earthquakes.*[5 marks]


#### 1.2: Loading Data \
The `.csv` data file given has been loaded onto a Google platform and published as
a URL in order to (hopefully) grant universal user access. However, this becomes defeatist when loading the following code chunk, which requires a redetermined
plot theme function. To ignore the error message, remove code chunk two. 

```{r Loading Data, inlcude=TRUE}
earthquake=read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vREAF1nhXX9kzvysMZQqlRIM6XPnlrjDtlmhXb8OsRcHuu2S3SIpI-xQnieAmIlQWfAeMP-IklnCZhl/pubhtml?gid=1389465964&single=true", col_names=FALSE) %>%
print()
```

```{r}
earth=read_sheet("https://docs.google.com/spreadsheets/d/e/2PACX-1vREAF1nhXX9kzvysMZQqlRIM6XPnlrjDtlmhXb8OsRcHuu2S3SIpI-xQnieAmIlQWfAeMP-IklnCZhl/pubhtml")
```



Okay... soooo. A *lot* of stuff happened, or tried to happen, here. I have removed
all lines of code which did not end up working out. But relevant thought processes 
that stuck around: \
- Working with the time kinda messed with my mind. The format given to us is 
like so: 1960-01-02T12:21:58.720Z (YYYY-MM-DD Time(military): HH:MM:SSS (TimeZone UTC). But the Zulu UTC time was not in the proper format (ex: - 0800) in order to use 
`%z` or `%Z` in any date formatting sequences.. which is why I removed that last 
part of the time column. We are not using the time change for any data analysis
here, which is why it got cut, but should have been needed we would have had to
continue figuring that out. 

```{r Loading Data Local 2, include=TRUE}
countries=read.csv(here("01_raw_data", "countries.csv"))  #load in datasets
earthquakes=read.csv(here("01_raw_data","earthquakes_5.5_1960-2023.csv")) %>%
  clean_names() %>%                                  #clean all col headers
  mutate(time=str_extract(time, "[[:graph:]T[:graph:]]*(?<=[:punct:])")) %>%
  mutate(time=as.POSIXct("1960-01-02T12:21:58.", format="%Y-%m-%dT%H:%M:%S.", 
                         tz="UTC")) %>%            #convert from chr to date
  mutate(country=str_extract(place, "(?<=[:graph:], )[:graph:]*")) %>% 
  mutate(country=remove.nas=TRUE) %>%
  arrange(desc(mag)) %>%                  #arrange by highest magnitude first
  select(time, country, everything()) %>% #reorder dt
  print()                                #View object created loads in properly 
```
Now that the dataset has been manipulated in the way that I want it, here are some observations that I notice: \
1. 

```{r}

```


#### 1.3: Loading Select Functions \

Lastly... We will be applying a pre-determined function that has previously been generated. Here, the function is titled theme_juicy.R, and will be used for any 
and all figures generated.
```{r Loading Functions, include=TRUE}
source(here("functions", "theme_juicy.R")) #load from 'functions' folder
theme_set(theme_juicy())                   #auto-apply as theme for plots
```




