---
title: "GG606 - Assignment 1"
author: "Student Name"
date: "`r paste(format(Sys.Date(), '%Y-%m-%d'))`"
output: html_document
---


## Assignment 1 - Importing, parsing, and querying data in the wild

1. Read the data in and clean it for analysis, using the readr package functions for reading and parsing data. Provide a few comments on the types of earthquakes and the sources of information of the earthquakes. \[5 marks\]

```{r echo=TRUE}
#code 
#***********************************
library(readr)


```

My answer is written here and is explains what I did and why.



2. Did more earthquakes happen on weekends or weekdays? Include a figure \[5 marks\]
```{r echo=TRUE}
#code 
#***********************************



```

The above analysis demonstrates...


3. Has there been any change in the frequency of earthquakes? Include a figure \[5 marks\]
```{r echo=TRUE}
#code 
#***********************************



```

The above analysis demonstrates...


4. Where were there more earthquakes in the 1990s, South America or North America? Include a figure \[5 marks\]
```{r echo=TRUE}
#code 
#***********************************



```

The above analysis demonstrates...


5. Comment on changes in the data with time: a) earthquakes per year, b) type of earthquakes, and c) magnitude resolution. Include a figure \[5 marks\]
```{r echo=TRUE}
#code 
#***********************************



```

The above analysis demonstrates...


6. Comment on how lessons from Wilke’s Fundamentals of Data Visualization were applied to each figure with specific reference to book sections \[5 marks\]

